---
title: "AssignmentCourse8"
author: "Saurabh Singh"
date: "7/5/2020"
output: html_document
---

## Summary
This document is the final report of the peer assessment project from
machine learning course, which is a part of data science specialization.
The purpose of this analysis is to predict the manner in which the six
participants performed the excercises described below and to answer the 
questions of associated course quiz. The machine leanring algorithm, which
uses the classe variable in training set, is applied to the 20 test cases 
availabe in the test data.

## Introduction
Devices such as Jawbone Up, Nike FuelBand, and Fitbit can enable collecting a 
large amount of data about someone’s physical activity. These devices are used 
by the enthusiasts who take measurements about themselves regularly to improve 
their health, to find patterns in their behavior, or because they are tech geeks. 
However, even though these enthusiasts regularly quantify how much of a 
particular activity they do, they rarely quantify how well they do it. In this 
project, the goal is to use data from accelerometers on the belt, forearm, arm, 
and dumbell of six participants. They were asked to perform barbell lifts 
correctly and incorrectly in five different ways.

```{r include=FALSE}
library(dplyr)
library(caret)
library(corrplot)
```

Load the training and test datasets
```{r loaddata}
data_train <- read.csv("pml-training.csv", na.strings = c("NA", ""))
data_quiz <- read.csv("pml-testing.csv", na.strings = c("NA", ""))
```

Create two partitions (75% and 25%) using the original training dataset
```{r partitions}
in_train  <- createDataPartition(data_train$classe, p=0.75, list=FALSE)
train_set <- data_train[ in_train, ]
test_set  <- data_train[-in_train, ]
```

The two datasets (train_set and test_set) have a large number of NA values variables.
columns with >20% NAs will be removed.
```{r removecols}
missingFrac <- colSums(is.na(train_set))/dim(train_set)[1]
selectCols <- names(missingFrac)[missingFrac<.2]
train_set <- train_set[ , selectCols]
test_set  <- test_set [ , selectCols]
```

Columns with near zero variation will also be removed from both the train and
the test datasets
```{r nzv}
nzv_var <- nearZeroVar(train_set)
train_set <- train_set[ , -nzv_var]
test_set  <- test_set [ , -nzv_var]
```

It appears that columns 1 to 5 are just for identification. Therefore,
these columns will also be removed
```{r firstfivecols}
train_set <- train_set[ , -(1:5)]
test_set  <- test_set [ , -(1:5)]
```

## Correlation Analysis
Correlation analysis is perromed to before the modeling work for exploratory
data analysis purpose. 
```{r corranalysis}
corr_matrix <- cor(train_set[ , -54])
corrplot(corr_matrix, order = "FPC", method = "circle", type = "lower",
         tl.cex = 0.6, tl.col = rgb(0, 0, 0))
```
There are only a few variables, which exhibit very high correlations. Note that
high correlation is indicated by dark blue or red colors. To reduce the number
of variables principal componenet analysis (PCA) can be performed. However, there
are only a few strong correlations among the input variables, we will not perform
PCA. 

## Prediction Models
### Linear Discriminant Analysis (LDA) Model
Train lda model on train_set, and predictions perfomed for test_set.
Confusion matrix was then generated using test_set.
```{r lda, cache=TRUE, message=FALSE, warning=FALSE}
set.seed(125)
modelFit_lda <- train(classe ~ ., method="lda", data=train_set)
pred_lda <- predict(modelFit_lda, newdata = test_set)
conf_matrix_lda <- confusionMatrix(pred_lda, test_set$classe)
conf_matrix_lda
```
The predictive accuracy of LDA model is relatively low at 71.7%

### Generalized Boosted Model (GBM)
```{r GBM1, cache=TRUE, message=FALSE, warning=FALSE}
set.seed(125)
ctrl_GBM <- trainControl(method = "repeatedcv", number = 5, repeats = 2)
fit_GBM  <- train(classe ~ ., data = train_set, method = "gbm",
                  trControl = ctrl_GBM, verbose = FALSE)
fit_GBM$finalModel
```
Predictions on the GBM test_set.
```{r, dependson="GBM1"}
predict_GBM <- predict(fit_GBM, newdata = test_set)
conf_matrix_GBM <- confusionMatrix(predict_GBM, test_set$classe)
conf_matrix_GBM
```
The predictive accuracy of the GBM is relatively high at 98.9%

### Random Forest Model
```{r RF1, cache=TRUE, message=FALSE, warning=FALSE}
set.seed(125)
ctrl_RF <- trainControl(method = "repeatedcv", number = 5, repeats = 2)
fit_RF  <- train(classe ~ ., data = train_set, method = "rf",
                  trControl = ctrl_RF, verbose = FALSE)
fit_RF$finalModel
```
Predictions of the random forest model on test_set.
```{r, dependson="RF1"}
predict_RF <- predict(fit_RF, newdata = test_set)
conf_matrix_RF <- confusionMatrix(predict_RF, test_set$classe)
conf_matrix_RF
```
The prediction accuracy of Random Forest model is 99.9%

### Applying the best predictive model (i.e. Random Forest) to the Test Data
The Random Forest model is selected and applied to make predictions on the
twenty data points from the original testing data set (data_quiz)
```{r quiz}
predict_quiz <- predict(fit_RF, data_quiz)
predict_quiz
```